---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dbmm: Dynamic Bayesian Measurement Models

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**dbmm** estimates latent attributes that evolve over time — the policy
liberalism of U.S. states, the ideological positions of demographic groups,
the human rights practices of countries — from indicators that need not
share a measurement scale. Models are written in
[Stan](https://mc-stan.org) and fit via
[**cmdstanr**](https://mc-stan.org/cmdstanr/).

The package implements two models:

- A **dynamic factor model for mixed-type indicators** (`*_mixfac()`), for
  units observed directly on indicators that may be binary, ordinal, or
  metric.
- A **multidimensional ordinal dynamic group-level IRT model**
  (`*_modgirt()`), for latent traits of groups inferred from the survey
  responses of their individual members.

Both models allow several latent dimensions, treat temporal dependence as a
random walk in the latent space, and supply post-estimation identification
restrictions — rotation, factor ordering, and sign normalization — along
with helpers for labelling, summarizing, and cross-validating the draws.

## Installation

**dbmm** requires a working CmdStan installation. If you do not have one:

``` r
install.packages("cmdstanr",
                 repos = c("https://mc-stan.org/r-packages/",
                           getOption("repos")))
cmdstanr::install_cmdstan()
```

Then:

``` r
# install.packages("pak")
pak::pak("devincaughey/dbmm")
```


## Quick start

The package ships with `state_policies_2010_2012`, a panel of 111 policies
of the 50 U.S. states observed annually from 2010 through 2012. Most
policies are binary indicators of whether a state had the policy in force;
others are ordinal, and some are continuous, such as tax rates and
per-pupil expenditures.

```{r eval=FALSE}
library(dbmm)

data("state_policies_2010_2012")

shaped <- shape_mixfac(
    long_data           = subset(state_policies_2010_2012, !is.na(value_real)),
    unit_var            = "state_abb",
    time_var            = "year",
    item_var            = "policy_variable",
    value_var           = "value_real",
    periods_to_estimate = 2010:2012,
    standardize         = TRUE
)
```

`shape_mixfac()` classifies each item by its number of distinct values —
here 64 binary, 24 trichotomous, 2 ordinal, and 18 metric items, with 3
dropped for lack of variation — and reports what it did. The
classification can be overridden with the `binary_items`,
`trichotomous_items`, and `ordinal_items` arguments.

```{r eval=FALSE}
fitted <- fit_mixfac(
    shaped_data     = shaped,
    n_dim           = 2,
    constant_alpha  = TRUE,
    smooth_eta      = TRUE,
    gen_log_lik     = TRUE,
    chains          = 4,
    parallel_chains = 4,
    iter_warmup     = 1000,
    iter_sampling   = 1000,
    seed            = 123
)
```

Arguments not listed in `?fit_mixfac` are passed through to **cmdstanr**'s
[`$sample()`](https://mc-stan.org/cmdstanr/reference/model-method-sample.html)
method.

When there are two or more latent dimensions, the likelihood is invariant
to rotation, reflection, and permutation of the factors, so the raw draws
are not directly interpretable. **dbmm** leaves the sampler unconstrained
and resolves the indeterminacy afterwards.

```{r eval=FALSE}
draws <- extract_mixfac_draws(fitted) |>
    identify_mixfac(method = "varimax", seed = 123) |>  # rotate and align
    label_mixfac() |>                                  # attach labels
    combine_types_mixfac() |>                          # stack item types
    sort_mixfac() |>                                   # order by variance
    sign_mixfac(signs = 1)                             # fix orientation

summarize_mixfac(draws)
```

Every one of these transformations is orthogonal or a relabelling, so the
linear predictors — and hence the likelihood — are left exactly unchanged.

Approximate leave-one-out cross-validation is available for models fit with
`gen_log_lik = TRUE`:

```{r eval=FALSE}
loo::loo_compare(
    one_factor = loo_mixfac(fitted_1d),
    two_factor = loo_mixfac(fitted_2d)
)
```

## Workflow

Both models follow the same sequence of steps.

| Step | Dynamic factor model | Group-level IRT model |
|------|----------------------|-----------------------|
| Shape data | `shape_mixfac()` | `shape_modgirt()` |
| Fit | `fit_mixfac()` | `fit_modgirt()` |
| Extract draws | `extract_mixfac_draws()` | *(taken from the fit)* |
| Identify | `identify_mixfac()` | `identify_modgirt()` |
| Label | `label_mixfac()` | `label_modgirt()` |
| Combine item types | `combine_types_mixfac()` | *(not needed)* |
| Order factors | `sort_mixfac()` | `sort_modgirt()` |
| Set signs | `sign_mixfac()` | `sign_modgirt()` |
| Summarize | `summarize_mixfac()` | `posterior::summarise_draws()` |
| Cross-validate | `loo_mixfac()` | — |

`identify_mixfac()` and `identify_modgirt()` apply a varimax rotation and a
signed permutation to each posterior draw; `rotate_modgirt()` applies a
rotation of your own choosing instead.

## Learn more

```{r eval=FALSE}
vignette("mixfac", package = "dbmm")   # dynamic factor model, mixed indicators
vignette("modgirt", package = "dbmm")  # group-level IRT from survey data
```

## Status

**dbmm** is experimental. The interface may change without a deprecation
cycle, and not every function is documented to the standard we would like.
Bug reports and questions are welcome at
<https://github.com/devincaughey/dbmm/issues>.

## Citation

```{r eval=FALSE}
citation("dbmm")
```

## License

GPL (>= 2)